---
output:
  html_document: default
  pdf_document: default
---

```{r knitr_setup, echo=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

# Joanne Chen
# CMSC320 Spring 2019 Final Project

# Honey Production in the USA from 1998-2012

Bees. Some of us may have had painful memories when it comes to bees. Apart from their stingers, 
bees are vital to Earth; for instance, plants need bees to pollinate. Pollination gives way to 
habitats for animals and food production for all. 

However, climate change, harmful pesticides, loss of habitat, etc. have significantly contributed
to the rapid decline of the honeybee population in the past decades. Subsequently, the honey
production has been negatively affected.

Data Science has helped researchers, environmentalists, and others observe the bee population
decline and bring more attention to the importance of taking actions towards protecting the bees.

Here we explore a dataset regarding the honey production in the United States of America from 
1998 to 2012, while also providing a walk through tutorial of the entire data science pipeline.

Sources:
  - https://www.earthday.org/campaigns/endangered-species/bees/
  - http://sos-bees.org/
  - https://www.planetbee.org/save-honeybees 

Dataset URL: https://www.kaggle.com/jessicali9530/honey-production 

## Pre-Processing

First, we are calling numerous libraries that are necessary for our code because these libraries 
consist of many useful R functions that will be used in the code.

```{r load_data, message=FALSE}
# Calling Libraries
library(tidyverse)
library(rvest)
library(dplyr)
library(ggplot2)
library(broom)
library(magrittr)
library(tidyr)
library(stringr)
```

## Data Loading and Parsing

After downloading a CSV file from Kaggle, use the read_csv() function with the file path as the 
parameter to read the data from a CSV file into a data frame in R. 
When loading a CSV file, you can parse it with column specifications using the cols() or 
cols_only() function. cols() includes all columns in the data, whereas cols_only() only include 
the columns specified.

For more information on the read_csv() function:   
  - https://readr.tidyverse.org/reference/read_delim.html 
for more information on the cols() and cols_only() function:
  - https://www.rdocumentation.org/packages/readr/versions/1.3.1/topics/cols 
Find datasets through links provided here: 
  - http://www.hcbravo.org/IntroDataSci/resources/ 
  - https://www.kaggle.com/datasets 
  
```{r}
df <- read_csv('/Users/JoanneChen/Desktop/honeyproduction.csv')

# Parsed with column specification
cols(
  state = col_character(),
  numcol = col_double(),
  yieldpercol = col_double(),
  totalprod = col_double(),
  stocks = col_double(),
  priceperlb = col_double(),
  prodvalue = col_double(),
  year = col_integer()
)
df
```

The set_colnames function allows for renaming column names.

```{r}
# Change column names
df <- df %>%
  set_colnames(c("state", "num_colonies", "yield_per_colony_lb", "total_production", "stocks", "price_per_lb", "production_value", "year"))
df
```

# Exploratory Data Analysis

Operations, such as select, filter, slice, arrange, group_by, and summarise, are used to help 
perform almost any analysis on data frames.

Using the various operations, we can determine which state has had the largest and smallest total
average honey production between 1998 and 2012, and on average, how many colonies it takes to 
create 1 pound of honey for each state.

Documentation of operations used:
  - group_by(): https://www.rdocumentation.org/packages/dplyr/versions/0.7.8/topics/group_by
  - summarise(): https://www.rdocumentation.org/packages/dplyr/versions/0.7.8/topics/summarise
  - filter(): https://www.rdocumentation.org/packages/dplyr/versions/0.7.8/topics/filter
  - mutate(): https://www.rdocumentation.org/packages/dplyr/versions/0.5.0/topics/mutate 
  
```{r}
# North Dakota (ND) has the largest average honey production
df %>%
  group_by(state) %>%
  summarise(avg_production = mean(total_production)) %>%
  filter(avg_production == max(avg_production))
```

```{r}
# Oklahoma (OK) has the smallest average honey production
df %>%
  group_by(state) %>%
  summarise(avg_production = mean(total_production)) %>%
  filter(avg_production == min(avg_production))
```

```{r}
# On average, the amount of colonies it takes to make 1 pound of honey for each state
df %>%
  group_by(state) %>%
  summarise(avg_production = mean(total_production), 
            avg_yield = mean(yield_per_colony_lb)) %>%
  mutate(num_colonies_1lb = avg_production/avg_yield)
```

The 4 plots below are all scatterplots of Total Honey Production in Pounds vs Year for all 
states. For majority of the states, the plots illustrate a gradual decrease in total honey 
production between the years 1998 and 2012.

geom_point() is used to create the scatterplot and geom_smooth() is used to create the trend 
line. labs() and ggtitle() is used to customize the names of the axises and title of the plot.
In addition, theme(plot.title = element_text(hjust = 0.5)) helps center the title of the plot.

Here is a ggplot2 reference sheet: https://ggplot2.tidyverse.org/reference/ 

```{r}
df %>%
  filter(state %in% c("CA", "FL", "ND", "MT", "MN", "SD", "TX")) %>%
  ggplot(aes(x = year, y = total_production, color = state)) + geom_point() + 
  geom_smooth(method = loess) + labs(x = "Year", y = "Total Yearly Honey Production (lb)") +
  ggtitle("Total Yearly Honey Production (lb) vs. Year") + 
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
df %>%
  filter(state %in% c("AZ", "GA", "IA", "UT", "PA", "CO", "AL", "HI")) %>%
  ggplot(aes(x = year, y = total_production, color = state)) + geom_point() + 
  geom_smooth(method = loess) + labs(x = "Year", y = "Total Yearly Honey Production (lb)") +
  ggtitle("Total Yearly Honey Production (lb) vs. Year") + 
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
df %>%
  filter(state %in% c("KS", "IL", "IN", "MD", "OK", "ME", "VA", "SC", "TN", "VT", 
                      "WV", "KY", "NM", "NJ")) %>%
  ggplot(aes(x = year, y = total_production, color = state)) + geom_point() + 
  geom_smooth(method = loess) + labs(x = "Year", y = "Total Yearly Honey Production (lb)") +
  ggtitle("Total Yearly Honey Production (lb) vs. Year") + 
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
df %>%
  filter(state %in% c("LA", "ID", "AR", "OR", "WI", "WY", "WA", "NY", "NE", "MI", 
                      "MS", "MO", "NC", "NV", "OH")) %>%
  ggplot(aes(x = year, y = total_production, color = state)) + geom_point() + 
  geom_smooth(method = loess) + labs(x = "Year", y = "Total Yearly Honey Production (lb)") +
  ggtitle("Total Yearly Honey Production (lb) vs. Year") + 
  theme(plot.title = element_text(hjust = 0.5))
``` 

Below is a scatterplot of the Average Total Honey Production in Pounds vs. Year.
geom_line() connects each point in order of the variable on the x-axis, highlighting the changes 
between each year in average honey production (lb).

Even though the total average honey production increases a little during 1999-2000, 2002-2004, 
2007-2008, an 2009-2010, as seen through the black line, overall, there is a general decrease in 
total average honey production from 1998 to 2012 as seen through the blue line.

```{r}
df %>%
  group_by(year) %>%
  summarise(avg_total_production = mean(total_production)) %>%
  ggplot(aes(x = year, y = avg_total_production)) + geom_point() + geom_line() +
  geom_smooth(method = loess) + 
  labs(x = "Year", y = "Average Total Yearly Honey Production (lb)") + 
  ggtitle("Average Yearly Total Honey Production (lb) vs. Year") + 
  theme(plot.title = element_text(hjust = 0.5))
``` 

Using geom_bar(), below is a bar graph of State vs. Total Honey Production in Pounds. The Total 
Honey Production number is the sum of all the honey a state has produced between 1998 and 2012. 
North Dakota (ND) has produced the most amount of honey, while South Carolina (SC) has produced 
the least amount of honey over time.

```{r}
df %>%
  group_by(state) %>%
  summarise(total_amount = sum(total_production)) %>%
  ggplot(mapping = aes(y = total_amount, x = reorder(state, total_amount), 
                       color = state, reorder(total_amount))) + 
  geom_bar(stat = "identity") + labs(y = "Total Honey Production (lb)", x = "State") + 
  ggtitle("Total Honey Production (lb) vs. State") + 
  theme(plot.title = element_text(hjust = 0.5)) + coord_flip()
``` 

Using geom_violin(), below is a violin plot of Price Per Pound ($) vs. Year. 
A violin plot is a type of plot that displays the distribution of the variable in the y-axis 
(price_per_lb) for each value of the variable in the x-axis (year).
Over the years, the price of honey per pound has gradually increased from less than 1 dollar to 
prices over 4 dollars.

```{r}
df %>%
  ggplot(aes(x = factor(year), y = price_per_lb)) + geom_violin() + 
  labs(x = "Year", y = "Price Per Pound ($)") + 
  ggtitle("Price Per Pound ($) vs. Year") + theme(plot.title = element_text(hjust = 0.5))
``` 

Below, using the cut() function, discretization is applied where year is split into 3 equal bins 
or intervals: (1998-2003], (2003-2007], (2007-2012]. These year intervals are then added into a 
new column, called year_interavls, in the data frame.

The scatterplot shows the Average Price Per Pound ($) vs. the Average Total Honey Production (lb)
in each year interval, using the facet_grid() function. From the 1st interval to the 3rd 
interval, there is an increasing trend in the average price of honey, while there is a decreasing
trend in the average total honey production. Florida (FL) and California (CA) are labeled to 
highlight this trend using the geom_text() function. The price increases approximately 50 cents 
per interval and the honey production decreases several million pounds per interval.

Function Documentation:
  - cut() ~ https://www.rdocumentation.org/packages/base/versions/3.6.0/topics/cut 
  - facet_grid() ~ https://ggplot2.tidyverse.org/reference/facet_grid.html
  - geom_text() ~ https://www.rdocumentation.org/packages/ggplot2/versions/0.9.1/topics/geom_text

```{r}
# Discretize year into 3 intervals
df$year_intervals <- cut(df$year, breaks = 3) 

df %>%
  group_by(state, year_intervals) %>%
  summarise(avg_production = mean(total_production), avg_price = mean(price_per_lb)) %>%
  ggplot(aes(x = avg_production, y = avg_price, color = state, label = state)) + geom_point() +
  geom_text(aes(label = ifelse(state == 'CA' | state == 'FL', state, '')), 
            hjust = 0, vjust = 0) + facet_grid(rows = vars(year_intervals)) + 
  labs(x = "Average Total Honey Production (lb)", y = "Average Price per Pound ($)") +
  ggtitle("Average Price per Pound ($) vs. Average Total Honey Production (lb)")
```  

Here is a dataframe of what was plotted above.

```{r}
df %>%
  group_by(state, year_intervals) %>%
  summarise(avg_production = mean(total_production), avg_price = mean(price_per_lb)) %>%
  arrange(state)
```  

# Linear Regression

Linear regression is a very useful technique for data analysis.
It allows for constructing confidence intervals, utilizing hypothesis testing for relationships 
between variables, and providing continuous outcomes of interest.

From http://r-statistics.co/Linear-Regression.html , 
  "linear regression is used to predict the value of an outcome variable Y based on one or more 
  predictor variables X. The aim is to establish a linear relationship (a mathematical formula) 
  between the predictor variable(s) and the response variable, so that, we can use this formula 
  to estimate the value of the response Y, when only the predictors (Xs) values are known."

For more information on Linear Regression and Linear Models:
  - https://www.statisticssolutions.com/what-is-linear-regression/
  - http://www.stat.yale.edu/Courses/1997-98/101/linreg.htm 
  - https://data.princeton.edu/r/linearmodels 
      - This link helps explains how to fit a model, examine a fit, extract results, and much   
      more.

```{r}
# Fit a linear regression model for Total Honey Production (lb) vs. Year
df_fit_honey <- lm(total_production~year, data = df)
df_fit_honey

df_fit_honey_stats <- df_fit_honey %>%
  tidy()
df_fit_honey_stats
``` 

```{r}
cat("On average, the total honey production decreased by", df_fit_honey_stats$estimate[2], 
    "pounds per year from 1998 to 2012.")
``` 

If there was a null hypothesis of no relationship between total honey production and year, I 
would reject that null hypothesis because there is a relationship between the two variables: Over
time, the total honey production decreases by around -88582.63 per year.

Now, we augment the df_fit_honey, in which columns, such as predictions, residuals, etc. are 
added.
Residuals is the difference between the observed value of the dependent variable and the 
predicted value. Fitted values are also known as predicted values.
One way to check if your linear regression model is appropriate is to plot a graph of Residuals 
vs. Fitted Values. This graph will check for the linearity assumption. If the regression model is
appropriate, the mean of residuals will be approximately 0.

The augment() function Documentation: 
  - https://www.rdocumentation.org/packages/broom/versions/0.4.3/topics/augment 
Residuals:
  - http://www.r-tutor.com/elementary-statistics/simple-linear-regression/residual-plot 
  
```{r}
aug_df_honey <- df_fit_honey %>%
  augment()

aug_df_honey
``` 

```{r}
# A violin plot of model Residuals vs. Year
aug_df_honey %>%
  ggplot(aes(x = factor(year), y = .resid)) + geom_violin() + 
  labs(x = "Year", y = "Residuals") + ggtitle("Residuals vs. Year") + 
  theme(plot.title = element_text(hjust = 0.5))
``` 

The 3 graphs below are violin plots of model Residuals vs. State. These 3 plots show that there 
is a dependence between model residuals and state because each violin is different. This suggests
that when performing a regression analysis of total honey production across time, it is important
to consider the state variable.

```{r}
aug_df_honey <- aug_df_honey %>%
  inner_join(df, by = "total_production")
aug_df_honey

aug_df_honey %>%
  filter(state %in% c("AL", "AR", "AZ", "CA", "CO", "FL", "GA","HI", "IA", 
                      "ID", "IL", "IN", "KS", "KY", "LA")) %>%
  ggplot(aes(x = factor(state), y = .resid)) + geom_violin() + 
  labs(x = "State", y = "Residuals") + ggtitle("Residuals vs. State") + 
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
aug_df_honey %>%
  filter(state %in% c("MD", "ME", "MI", "MN", "MO", "MS", "MT", "NC", 
                      "ND", "NE", "NJ", "NM", "NV", "NY", "OH")) %>%
  ggplot(aes(x = factor(state), y = .resid)) + geom_violin() + 
  labs(x = "State", y = "Residuals") + ggtitle("Residuals vs. State") + 
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
aug_df_honey %>%
  filter(state %in% c("OK", "OR", "PA", "SC", "SD", "TN", "TX", "UT", 
                      "VA", "VT", "WA", "WI", "WV", "WY")) %>%
  ggplot(aes(x = factor(state), y = .resid)) + geom_violin() + 
  labs(x = "State", y = "Residuals") + ggtitle("Residuals vs. State") + 
  theme(plot.title = element_text(hjust = 0.5))
```

Here, we fit a linear regression model for total honey production including a term for an 
interaction between state and year.

```{r}
state_honey_fit <- lm(total_production~year*state, data = df)
state_honey_fit

state_honey_fit_stats <- state_honey_fit %>%
  tidy()
state_honey_fit_stats
```

The state_honey_fit_stats allows us to find out how much total honey production (in pounds) 
increases or decreases each year on average for each state. For instance, below is how you would 
calculate how much total honey production changes per year for AL, HI, and CA.

```{r}
cat("AL:", state_honey_fit_stats$estimate[2], "\n")
cat("HI:", (state_honey_fit_stats$estimate[2] + state_honey_fit_stats$estimate[9]), "\n")
cat("CA:", (state_honey_fit_stats$estimate[2] + state_honey_fit_stats$estimate[5]))
```

Below, we perform a F-test that compares the Linear Regression with and without the state 
variable using the anova() function.
The anova() function Documentation:
  -https://www.rdocumentation.org/packages/car/versions/3.0-2/topics/Anova 

```{r}
comp_honey <- anova(df_fit_honey, state_honey_fit)
comp_honey
```

The interaction model (or Model #2) is significantly better than the year-only model because the 
p-value for Model #2 is ver low at 2.2e-16, meaning that adding state to the model significantly 
improved fit over Model #1

Here are some links for more information on how to interpret:
  - https://stats.stackexchange.com/questions/115304/interpreting-output-from-anova-when-using-lm-as-input 
  - https://bookdown.org/ndphillips/YaRrr/comparing-regression-models-with-anova.html 

Below, we'll do the same but with the price of honey per pound.

```{r}
# Fit a linear regression model for Price of Honey per Pound ($) vs. Year
df_fit_price <- lm(price_per_lb~year, data = df)
df_fit_price

df_fit_price_stats <- df_fit_price %>%
  tidy()
df_fit_price_stats
``` 

```{r}
cat("On average, the price per pound of honey increased by", df_fit_price_stats$estimate[2],
    "cents per year from 1998 to 2012.")
```

If there was a null hypothesis of no relationship between the price of honey per pound and year, 
I would reject that null hypothesis because there is a relationship between the two variables: 
Over time, the price of honey per pound increases by approximately $0.10.

```{r}
aug_df_price <- df_fit_price %>%
  augment()
aug_df_price
```

```{r}
# A violin plot of model Residuals vs. Year
aug_df_price %>%
  ggplot(aes(x = factor(year), y = .resid)) + geom_violin() + 
  labs(x = "Year", y = "Residuals") + ggtitle("Residuals vs. Year") + 
  theme(plot.title = element_text(hjust = 0.5))
``` 

The 3 graphs below are violin plots of model Residuals vs. State. These 3 plots show that there 
is a dependence between model residuals and state because each violin is different. This suggests
that when performing a regression analysis of price of honey per pound across time, it is 
important to consider the state variable.

```{r}
aug_df_price <- aug_df_price %>%
  inner_join(df, by = "price_per_lb")
aug_df_price

aug_df_price %>%
  filter(state %in% c("AL", "AR", "AZ", "CA", "CO", "FL", "GA","HI", "IA", 
                      "ID", "IL", "IN", "KS", "KY", "LA")) %>%
  ggplot(aes(x = factor(state), y = .resid)) + geom_violin() + 
  labs(x = "State", y = "Residuals") + ggtitle("Residuals vs. State") + 
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
aug_df_price %>%
  filter(state %in% c("MD", "ME", "MI", "MN", "MO", "MS", "MT", "NC", 
                      "ND", "NE", "NJ", "NM", "NV", "NY", "OH")) %>%
  ggplot(aes(x = factor(state), y = .resid)) + geom_violin() + 
  labs(x = "State", y = "Residuals") + ggtitle("Residuals vs. State") + 
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
aug_df_price %>%
  filter(state %in% c("OK", "OR", "PA", "SC", "SD", "TN", "TX", "UT", 
                      "VA", "VT", "WA", "WI", "WV", "WY")) %>%
  ggplot(aes(x = factor(state), y = .resid)) + geom_violin() + 
  labs(x = "State", y = "Residuals") + ggtitle("Residuals vs. State") + 
  theme(plot.title = element_text(hjust = 0.5))
```

Here, we fit a linear regression model for price of honey per pound including a term for an 
interaction between state and year.

```{r}
state_price_fit <- lm(price_per_lb~year+state, data=df)
state_price_fit

state_price_fit_stats <- state_price_fit %>%
  tidy()
state_price_fit_stats
```

The state_price_fit_stats allows us to find out how much price per pound of honey increases or 
decreases each year on average for each state. For instance, below is how you would calculate how
much total honey production changes per year for AL, VA, and TN.

```{r}
cat("AL:", state_price_fit_stats$estimate[2], "\n")
cat("VA:", (state_price_fit_stats$estimate[2] + state_price_fit_stats$estimate[40]), "\n")
cat("TN:", (state_price_fit_stats$estimate[2] + state_price_fit_stats$estimate[37]))
```

Below, we perform a F-test that compares the Linear Regression with and without the state 
variable using the anova() function.

```{r}
comp_price <- anova(df_fit_price, state_price_fit)
comp_price
```

The interaction model (or Model #2) is significantly better than the year-only model because the 
p-value for Model #2 is ver low at 2.2e-16, meaning that adding state to the model significantly 
improved fit over Model #1

# Conclusion

Through analyzing the data from the National Agricultural Statistics Service (NASS) of the U.S. 
Department of Agriculture (USDA), it is evident that in almost all 44 out of 50 states included 
in this dataset, the honey production has decreased from 1998-2012, subsequently causing the 
price of honey to increase. 

According to the NASS and USDA, the honey Americans consume per year now mostly comes from 
foreign countries out of the US, instead of locally. The decline of honey production is largely 
due to the decline of the bee population.

Actions YOU can take to protect the bees and the honey!!!
  ~ Educate yourself
  ~ Plant a pollinator garden
  ~ Stop using pesticides
  ~ Get involved in various projects:
     - https://www.greatsunflower.org/ 
     - https://www.planetbee.org/zbw
     - https://honeybeenet.gsfc.nasa.gov/Sites.htm 
  ~ Support beekeepers
  ~ Become a bee keeper :D 
